<!-- R Commander Markdown Template -->

MAPOR 2017 Coding Hour
======================

### Trent Buskirk and Stas Kolenikov

### 2017-08-28

# Script 1: data import

First, let us navigate to the directory where the data are sitting -- this is specific to your computer.
This is done with `setwd()`, set working directory.



```{r}
setwd("C:/Users/kolenikovs/Conferences/MAPOR.2017/R with Trent")
```

## Import from Excel

### Example 1.1

Flexible and powerful import from Excel is provided by package `library(readxl)`:



```{r}
library(readxl)
Rdata1 <- readXL("R4SPSSData.xlsx", 
	rownames=FALSE, header=TRUE, na="", sheet="Sheet1", stringsAsFactors=TRUE)
```


Note how R Commander GUI creates a full R command and sends it to R. In the chunk of code
above, all options are explicitly spelled out.

## Categorical variables 

### Example 1.2

Unlike SPSS or Stata, R does not have a concept of value labels (SAS people may refer to them as formats).
To designate a variable as a categorical one to R, use `base::factor()` or `base::as.factor()` function.
(Here, we are using a standard R notation of `package::function()`, to make it explicit where each
function is to be found, and which `library(package)` may need to be loaded to make that function available.
The `base` package does not need to be loaded specifically; the base functions are always available in R.)
The former allows explicit mapping of `levels` of the original variables into `labels` of the factor
variable (in the order the values are supplied in both options). 



```{r}
Rdata1$Internet <- factor(Rdata1$InternetHH, levels=c(0,1), labels=c("NO","YES"))
table(Rdata1$Internet, Rdata1$InternetHH)
```

```{r}
Rdata1$Own <- factor(Rdata1$OwnRent, levels=c(1,2), labels=c("Own","Rent"))
table(Rdata1$OwnRent, Rdata1$Own)
```


Creating percentile groups of data requires a bit more work. We will first use function `base::rank()`
to determine the rank of each value in the full data (`base::cbind()` is used to put columns side by side
for demonstration purposes):

```{r}
cbind(Rdata1$TotTWLSW,rank(Rdata1$TotTWLSW))
```

We will then convert this to cdf, of sorts, by dividing the `rank()` by the number of observations:


```{r}
cbind(Rdata1$TotTWLSW,rank(Rdata1$TotTWLSW)/length(Rdata1$TotTWLSW))
```


Finally, to convert to quartile groups (0-25%-50%-75%-100%), we need to multiply by 4 and round 
to the next greatest integer (`base::ceiling()` function):

```{r}
Rdata1$TotTWQ <- ceiling(4*rank(Rdata1$TotTWLSW)/length(Rdata1$TotTWLSW))
cbind(Rdata1$TotTWLSW, Rdata1$TotTWQ )
```

The final bit of cleaning is to remove the data entry with missing `Hhsize`. This is done 
by indexing the data set with the square bracket notation `[,]`. The first argument of 
the square brackets is the row or set of rows; the second argument is the column or a set
of columns. The function `base:is.na()` evaluates whether its argument is missing `NA` or not,
and the exclamation mark is the logical negation. Thus `is.na(Rdata1$Hhsize)` identifies
the row(s) that have missing values `NA` of the variable `Hhsize`, negate that (so that only
the rows with nonmissing values are selected), and these rows are put as the argument
for subsetting. The second argument, the columns, is empty, which is a convention to 
select everything available.


```{r}
Rdata1nm <- Rdata1[ !is.na(Rdata1$Hhsize), ]
Rdata1nm
```

To rename the variable `Hhsize`, a little bit of effort is required, as R does not have
a simple function for that:

```{r}
colnames(Rdata1nm)
colnames(Rdata1nm)[colnames(Rdata1nm)=='Hhsize'] <- 'HHsizeNM'
colnames(Rdata1nm)
```

## Data merge

### Example 1.3

This example requires importing an SPSS file. The R package that handles this is `library(foreign)`.
It provides, amon others, the function `foreign::read.spss()` for the purpose. By default,
R Commander automatically converts SPSS names to lowercase, which makes sense in the context
of R conventions so that VARIABLE NAMES DO NOT SHOUT. We turned off the default to convert 
labeled SPSS variables into R factor variables, as specified in option `use.value.labels=FALSE`.

Part (A) code and output are:

```{r}
library(foreign)
Data2 <- read.spss("R4SPSSDataNumber2.sav", 
  use.value.labels=FALSE, max.value.labels=Inf, to.data.frame=TRUE)
colnames(Data2) <- tolower(colnames(Data2))
head(Data2)
```

In part (B), we need to load an existing R data set, which should be a piece of cake,
and rename one of the variables to lower case. The standard R function for the former
is `base::load()`, and we've done renaming before. Here, we use a string function `base::tolower()`
to make it explicit what it is exactly that we are doing. Unlike SAS or SPSS, the case
matters in R.


```{r}
load("Data3.Rda")
colnames(Data3)[colnames(Data3)=='ID'] <- tolower('ID')
head(Data3)
```

In part (C), we are going to merge two data frames (not data files, as you would probably
be used to in SPSS, SAS or Stata; R holds objects in memory, and the base `base::merge()` function 
applies specifically on data frames). R merges two data frames at a time, and if there are 
variable name conflicts, R will staple suffixes like `.x` and `.y` to distinguish the variables
that came from one or the other data set.


```{r}
R4SPSS2 <- merge(Data2, Data3, by='id')
head(R4SPSS2)
```

In part (D), we are going to standardize the income variable in `R4SPSS2`, 
(calling it `IncStand`), and delete the `HHIncome` variable in the data file.   
Being done in R Commander, it spits a bunch of code that can be simplified into one line
(`cbind()` is only used to display the original and scaled variables together):


```{r}
R4SPSS2$IncStand <- scale(R4SPSS2$HHIncome)
cbind( R4SPSS2$IncStand, R4SPSS2$HHIncome)
```

Finally, to drop a variable from a data frame, we can use `base::subset()`. The minus sign
in front of the variable name in `select` option indicates that this variable needs
to be anti-selected, i.e. *removed*. Note also that the variable name should go
without quotes.

```{r}
R4SPSS2 <- subset(R4SPSS2, select = -HHIncome )
head(R4SPSS2)
```

In part (E), we will save the final data file as an R data file named `R4SPSS2`,
which is done via a very reasonably named `base::save()` function:


```{r}
save(R4SPSS2,file='R4SPSS2.Rda')
```

We are done with data entering and transformations for this part of the course!
Please proceed to Script 2.

