<!-- R Commander Markdown Template -->

MAPOR 2017 Coding Hour: Graphics
================================

### Trent Buskirk and Stas Kolenikov

### `r as.character(Sys.Date())`

```{r echo=FALSE}
# include this code chunk as-is to set options
knitr::opts_chunk$set(comment=NA, prompt=TRUE, out.width=750, fig.height=8, fig.width=8)
library(Rcmdr)
library(car)
library(RcmdrMisc)
```


```{r echo=FALSE}
# include this code chunk as-is to enable 3D graphs
library(rgl)
knitr::knit_hooks$set(webgl = hook_webgl)
```


## MAPOR 2017 R coding hour (Buskirk + Kolenikov)

## Script 2: graphics

### Index plot

```{r}
setwd("C:/Users/kolenikovs/Conferences/MAPOR.2017/R with Trent")
load("datasets/WorkshopData.RData")
Rtest <- rtest
with(Rtest, indexplot(NumTVs, type='h', id.method='y', id.n=2, 
  labels=rownames(Rtest)))
```

### Histogram

This is a default run:

```{r}
with(Rtest, Hist(Hhsize, scale="frequency", breaks="Sturges", 
  col="darkgray"))
```

First, in R Commander GUI, we selected the `percent` option as the y-axis:

```{r}
with(Rtest, Hist(Hhsize, scale="percent", breaks="Sturges", col="darkgray", 
  xlab="Household size"))
```

Then we changed the color to green (`#00CD00`)...

```{r}
with(Rtest, Hist(Hhsize, scale="percent", breaks="Sturges", col=3, 
  xlab="Household size"))
```

... and to blue (`#0000FF`)

```{r}
with(Rtest, Hist(Hhsize, scale="percent", breaks="Sturges", col=4, 
  xlab="Household size"))
```

Export the graph file to disk:

```{r}
dev.print(png, 
+   filename="HHsize-hist.png",
+    width=1000, height=998, pointsize=12, units="px")
```

### Kernel density plot

Kernel density estimate is a version of density representation
that produces a smooth line rather than bars, like historgrams do.

```{r}
densityPlot( ~ AgeofPanelist, data=Rtest, bw="SJ", adjust=1, 
  kernel="epanechnikov", xlab="Age")
```

### Scatterplots

Scatterplots demonstrate the relations between two continuous variables.
The relations may be monotone (i.e., one variable goes up as the other
goes up or down), curvlinear with change of trends, cluster (i.e.,
groups of observations separated in space), etc.

```{r}
scatterplot(TotTWLSW~AgeofPanelist, reg.line=FALSE, smooth=TRUE, 
  spread=FALSE, boxplots='xy', span=0.5, ellipse=FALSE, levels=c(.5, .9), 
  data=Rtest)
```

### Plotting symbols and linetypes (slide 66)

Here are a couple of "classic" plots showing R symbols and line types.

### Violin plot

```{r}
library(vioplot)
vioplot(Rtest$AgeofPanelist,col="green3")
```

## Example 2.1

> Using the dataBodyMeas dataset in the lessR package:

> (a) plot a histogram of heights and use percentages as the basis of histogram (as opposed to frequencies)

> (b) plot a density estimate of the heights (adjust =1.2)

> (c) generate a boxplot of the heights.

> (d) generate a violin plot of the heights – try two different values of the `h` parameter – try .3 and 1 and 4 
and three different colors = "green", "red" and "magenta"

```{r}
library(lessR)
data(dataBodyMeas)
```

### Example 2.1.a

> plot a histogram of heights and use percentages as the basis of histogram (as opposed to frequencies)

```{r}
Hist(dataBodyMeas$Height,scale="percent")
```

### Example 2.1.b

> plot a density estimate of the heights (adjust =1.2)

```{r}
densityPlot( ~ Height, data=dataBodyMeas, bw="SJ", adjust=1.2, 
  kernel="epanechnikov")
```

### Example 2.1.c

> generate a boxplot of the heights.

```{r}
Boxplot( ~ Height, data=dataBodyMeas, id.method="y")
```

### Example 2.1.d

> generate a violin plot of the heights – try two different values of the `h` parameter – try .3 and 1 and 4 
and three different colors = "green", "red" and "magenta"


```{r}
vioplot(dataBodyMeas$Height, col=c("green"), h=.3, horizontal=TRUE)
title("Violin plot of Heights; h=.3", cex.main=1.5)
```

```{r}
vioplot(dataBodyMeas$Height, col=c("red"), h=1, horizontal=TRUE)
title("Violin plot of Heights; h=1", cex.main=1.5)
```

```{r}
vioplot(dataBodyMeas$Height, col=c("magenta"), h=4, horizontal=TRUE)
title("Violin plot of Heights; h=4", cex.main=1.5)
```

## Example 2.2

### Example 2.2.a 

> Using dataBodyMeas dataset in the lessR package, plot a density estimate of the heights distribution 
for males and females using the Epanechnikov kernel function with default values. (multiplier of 1)

```{r}
densityPlot(Height~Gender, data=dataBodyMeas, bw="SJ", adjust=1, 
   kernel="epanechnikov")
```

### Example 2.2.b

> Repeat (a) but apply a multiplier of 1 for females and 5 for males 
to the default bandwidth.  [Males have larger variability so we increase 
the multiplier to “smooth out the plot”].  Use the 2nd and 4th colors 
in your palette.  

```{r}
densityPlot(Height~Gender, data=dataBodyMeas, bw="SJ", adjust=c(1,5), 
  kernel="epanechnikov",col=c(2,4))
```

### Example 2.2.c

> Create side-by-side boxplots of the distribution of the weight by sex.  
Use the third color in your color palette for this plot.  

```{r}
Boxplot(Weight~Gender,data=dataBodyMeas,col=2,id.method="none")
```

## Scatterplots and scatterplot matrices

Rcommander can generate scatterplots that have extended features including:
* loess smothed curves (locally weighted scatterplot smoothing)
* spread of the smooth curve (variance+bias)
* least squares regression line
* marginal boxplots

Rcommander can also generate scatterplot matrices
* histograms, density plots or boxplots along the diagonal


```{r}
library(lattice, pos=16)
xyplot(Hips + Waist ~ Height | Gender, outer=TRUE, layout=c( 2 * 1 , 2 ) , 
  between=list(x=c( 0 ), y=1), type="p", pch=16, auto.key=list(border=TRUE), 
  par.settings=simpleTheme(pch=16), scales=list(x=list(relation='same'), 
  y=list(relation='same')), data=dataBodyMeas)
```


## Example 2.3

Using `data(dataBodyMeas)`, produce a bunch of scatterplots

### Example 2.3.a

> Using the dataBodyMeas dataset in the lessR package, create a scatterplot 
depicting the relationship between waist (X) and hip circumference (Y).  

> *  Label the x-axis “Waist Circumference” and the y-axis “Hip Measure”.  
Also plot marginal boxplots on the scatterplot axes.  
All labels should have a multiplier of 1.2.    Point size should also have a multiplier of 1.3 and NO points should be labelled. 

```{r}
scatterplot(Hips~Waist, reg.line=FALSE, smooth=FALSE, spread=FALSE, 
  boxplots='xy', span=0.5, ellipse=FALSE, levels=c(.5, .9), xlab="Waist 
  Circumference", ylab="Hip Measure", cex=1.3, cex.axis=1.2, cex.lab=1.2, 
  data=dataBodyMeas)
```

As cases with identical hip/waist measurements are glued together on the plot,
it may make sense to jitter both measures, so that the said overlap is broken.

```{r}
scatterplot(Hips~Waist, reg.line=FALSE, smooth=FALSE, spread=FALSE, 
  boxplots='xy', span=0.5, ellipse=FALSE, levels=c(.5, .9), jitter=list(x=1, 
  y=1), xlab="Waist Circumference", ylab="Hip Measure", cex=1.3, cex.axis=1.2,
   cex.lab=1.2, data=dataBodyMeas)
```

### Example 2.3.b

> Using this same dataset, create a scatterplot matrix of height, weight and 
waist size.
> * choose histograms for the diagonal and do not plot additional lines or 
identify any points within each plot.


```{r}
scatterplotMatrix(~Height+Waist+Weight, reg.line=FALSE, smooth=FALSE, 
  spread=FALSE, span=0.5, ellipse=FALSE, levels=c(.5, .9), id.n=0, diagonal = 
  'histogram', data=dataBodyMeas)
```

### Example 2.3.c

> Create a `ShoeBin` variable that has 4 groups of equal size, 
then use this variable to construct a plot of mean heights for each of the four 
`Shoebin` groups.  Include standard deviations in the plot.  
> * Label the y-axis "Mean Heights (and Std. Dev)" and the x-axis 
"Shoe Size Quartile Groups"

Creating four bins of `Shoe`:

```{r}
dataBodyMeas$ShoeBin4 <- with(dataBodyMeas, bin.var(Shoe, bins=4, 
  method='proportions', labels=NULL))
table(dataBodyMeas$ShoeBin4)
```

Plotting height by quartiles of `ShoeBin4`:


```{r}
with(dataBodyMeas, plotMeans(Height, ShoeBin4, error.bars="sd", 
  xlab="Shoe Size Quartile Groups", ylab="Mean Heights and St Devs", 
  connect=TRUE))
```

### Example 2.3.d

> Create a plot of `height` (X) versus `weight` (Y) by conditioning on 
`shoe` size quartile and `gender`.  
> * Allow the x-axis  to remain the same for all plots, 
but allow the y-axis to vary freely.  Use the plotting symbol of 
a solid square (`pch=15`) and make its color the fourth one in your palette.  


```{r}
xyplot(Weight ~ Height | Gender + ShoeBin4, type="p", pch=15, 
  auto.key=list(border=TRUE), par.settings=simpleTheme(pch=15), 
  scales=list(x=list(relation='same'), y=list(relation='free')), 
  data=dataBodyMeas, col=4)
```

## Plots for categorical variables

`Rcommander` can easily produce **basic** graphs specifically relevant for categorical variables including:
* Pie Graphs
* Bar Charts (best with one categorical variable)
`lessR` graphics capabilities for categorical variables extends the capabilities of Rcommander 
* Bubble plots
* expanded bar charts (multiple categorical variables)
* Sunflower plots 

### Bar charts

From RCommander GUI:

```{r}
with(dataBodyMeas, Barplot(ShoeBin4, xlab="ShoeBin4", ylab="Frequency"))
```

Using `lessR`:

```{r}
 BarChart(ShoeBin4, data=dataBodyMeas, horiz=TRUE)
```

Note the interesting suggestions `lessR` made:

```{r message=FALSE}
Plot(ShoeBin4, data=dataBodyMeas)  # bubble plot
Plot(ShoeBin4, values="count", data=dataBodyMeas)  # scatter plot
BarChart(ShoeBin4, horiz=TRUE, data=dataBodyMeas)  # horizontal bar chart
BarChart(ShoeBin4, colors="rainbow", data=dataBodyMeas)  # different bar colors 
```

Note, however, that `lessR` sometimes fails to pick up the right variable,
and plots `Gender` instead of `ShoeBin4`.

Some suggested variations:

```{r}
dataBodyMeas$ShoeBin4 <- with(dataBodyMeas, bin.var(Shoe, bins=4, 
  method='proportions', labels=NULL))
# this has to be forced, as R reloads dataBodyMeas, apparently
BarChart(Gender, by=ShoeBin4, data=dataBodyMeas)
BarChart(Gender, by=ShoeBin4, data=dataBodyMeas, horiz=TRUE)
BarChart(Gender, by=ShoeBin4, data=dataBodyMeas, beside=TRUE)
BarChart(Gender, by=ShoeBin4, data=dataBodyMeas, prop=TRUE)
```

### Bubble plots

Bubble plots enhance two or three-dimensional scatterplots by allowing 
the inclusion of an additional variable that governs the size of plotted points.
Bubble plots can be drawn easily using the `ScatterPlot()` function in the `lessR` 
package.

```{r}
ScatterPlot(Height, Weight, data=dataBodyMeas, size=Waist, 
bubble.power=.25, shape="circle")
```

Note that some of the `bubble` options have apparently changed
between versions.

Bubble plots can be conveniently used to visualize categorical
variables, as well. Bubble plots applied to two categorical variables 
(stored as numeric variables in R with fewer than 10 unique values each) 
is a way to create a more informative scatterplot where the size of 
the plotted point is proportional to another value. 
Commonly this value is the frequency of  the cell formed by the 
cross tabulation of the two categorical variables.

```{r}
mydata <- dataBodyMeas
mydata$ShoeBin4 <- with(mydata, bin.var(Shoe, bins=4, 
  method='proportions', labels=NULL))
dataBodyMeasSum <- as.data.frame(table(mydata$ShoeBin4,mydata$Gender))
colnames(dataBodyMeasSum) <- c("ShoeBin4","Gender","Freq")
ScatterPlot(Gender,ShoeBin4,data=dataBodyMeasSum,size=Freq)
```

Another variation on the topic of using a count variable
to be added to the location on the scatterplot is the sunflower plot.
In this plot, each point is depicted as a "sunflower" with the number 
of petals being proportional to another variable -- most commonly, 
the joint frequency of the specific levels of the two 
categorical variables, considered together in a cross-tabulated 
contingency table.

```{r}
# canonical version -- not sure if it works
ScatterPlot(Gender,ShoeBin4,data=mydata,size=Freq,shape="sunflower")
# alternative -- the ticks and axes work poorly though
sunflowerplot(ShoeBin4~Gender,data=dataBodyMeas)
sunflowerplot(ShoeBin4~Gender,data=dataBodyMeas,xlim=c(0.5,2.5),ylim=c(0.5,4.5),axes=FALSE)
axis(side=1,at=c(1,2),labels=c("Female","Male"),lwd=0)
axis(side=2,at=1:4,lwd=0)
```

## Example 2.4

The dataset `dataMach4` from the `lessR` package contains `Gender` and 
20 Likert scale items assessing the degree of Machiavellianism present 
in respondents. The Likert scales are 0=Strongly Disagree; 5=Strongly Agree.

```{r}
data(dataMach4)
dataMach4 <- within(dataMach4, {
  GenderNum <- as.factor(Gender)
})
LikertCats <- c("Strongly Disagree", "Disagree", "Slightly Disagree","Slightly Agree", "Agree", "Strongly Agree")
```

### Example 2.4.b

> Create a horizontal bar chart for item `m06`: 
> * Honesty is the Best Policy 
> displaying proportions instead of frequencies


```{r}
BarChart(m06,data=dataMach4,horiz=TRUE,prop=TRUE)
```

### Example 2.4.c

> Create a vertical bar chart for item `m07`: No Excuse for lying 
> * separated by `Gender` 

```{r}
BarChart(Gender,data=dataMach4,by=m07,beside=TRUE)
```

The color scheme is different, and the variable label did not get through.

### Example 2.4.d

```{r}
ScatterPlot(m06,  m07, data=dataMach4, value.labels=LikertCats, shape="sunflower")
```

# The End

That's all folks!

