<!-- R Commander Markdown Template -->

MAPOR 2017 Coding Hour: Setup
=============================

### Trent Buskirk and Stas Kolenikov

### `r as.character(Sys.Date())`

```{r echo=FALSE}
# include this code chunk as-is to set options
knitr::opts_chunk$set(comment=NA, prompt=TRUE, out.width=750, fig.height=8, fig.width=8)
library(Rcmdr)
library(car)
library(RcmdrMisc)
```


```{r echo=FALSE}
# include this code chunk as-is to enable 3D graphs
library(rgl)
knitr::knit_hooks$set(webgl = hook_webgl)
```


## MAPOR 2017 R coding hour (Buskirk + Kolenikov)

## Script 0: setup

This is the list of recommended packages for this training:
```{r}
what.we.need <- c(
	"Rcmdr", "lessR", "MASS", "boot", "nlme", "randomForest", "foreign",
	"lattice", "MatchIt", "ggplot2", "survey", "PASWR", "readxl"
)
```

Let us first figure out which packages you already have installed:

```{r}
packages <- rownames(installed.packages())
```

It looks like there are `r length(packages)` packages installed in the system already.

Let's see if there are any packages that your system is missing:
```{r collapse=TRUE}
to_install <- what.we.need[!(what.we.need %in% packages)]
to_install
```

If there is anything missing (i.e., the vector `to_install` has at least one entry,
which is what we check with `if (length(to_install) > 0)`, we can install it now
(you can expect a window to pop-up and ask you to select the CRAN mirror for this session):

```{r collapse=TRUE}
if (length(to_install) > 0) {
  install.packages(to_install)
}
```

Now, load all the packages that we need (output will be suppressed).
The function `lapply` goes over the list supplied as the first argument,
and applies the function that is its second argument. Everything else
(e.g., in this case, the option `character.only = TRUE`) is passed as 
parameter(s) of that function. 

```{r message=FALSE, results="hide"}
lapply(what.we.need, library, character.only = TRUE)
```

You should be all set now! Please proceed to Script 1.

